Negative and Fractional Types

### 2012-06-28 rational_ext.tex
small tweaks in section 2 to make things more precise

For example {{ swap* }} induces a function 
%subcode{code}
%! language = Haskell
% swapProd (a,b) = (b,a)
The isomorphisms are extended to form a congruence relation by adding the
following constructors that witness equivalence and compatible closure:
%subcode{proof} include main
%@  ~
%@@ id : b <-> b 
%
%@ c : b1 <-> b2
%@@ sym c : b2 <-> b1
%
%@ c1 : b1 <-> b2
%@ c2 : b2 <-> b3
%@@ c1(;)c2 : b1 <-> b3
%---
%@ c1 : b1 <-> b3
%@ c2 : b2 <-> b4
%@@ c1 (+) c2 : b1 + b2 <-> b3 + b4
%
%@ c1 : b1 <-> b3
%@ c2 : b2 <-> b4
%@@ c1 (*) c2 : b1 * b2 <-> b3 * b4

@@

@@

\begin{definition}{(Syntax of {{Pi}})}
\label{def:Pi}
We collect our types, values, and isomorphisms, to get the full language
definition.
%subcode{bnf} include main
% value types, b ::= 0 | 1 | b+b | b*b 
% values, v ::= () | left v | right v | (v,v) 
%
% iso.~types, t ::= b <-> b
% base iso ::= zeroe | zeroi 
%     &|& swap+ | assocl+ | assocr+ 
%     &|& identl* | identr* 
%     &|& swap* | assocl* | assocr* 
%     &|& dist0 | factor0 | dist | factor 
% iso comb., c ::= iso | id | sym c | c (;) c | c (+) c | c (*) c 
\end{definition}

@@

The operational semantics of {{Pi}} is summarized below (also see
\cite{infeffects}).  The semantics of the primitive combinators is
given by the following single-step reductions below. Since there are
no values of type {{0}}, there are no rules for the impossible cases:
\begin{scriptsize}
%subcode{opsem} include main
%! columnStyle = rlcl
% zeroi & v & |-->& right v
% zeroe & (right v) &|-->& v
% swap+ & (left v) &|-->& right v
% swap+ & (right v) &|-->& left v 
% assocl+ & (left v1) &|-->& left (left v1)
% assocl+ & (right (left v2)) &|-->& left (right v2)
% assocl+ & (right (right v3)) &|-->& right v3 
% assocr+ & (left (left v1)) &|-->& left v1
% assocr+ & (left (right v2)) &|-->& right (left v2)
% assocr+ & (right v3) &|-->& right (right v3)
% identl* & ((), v) &|-->& v 
% identr* & v &|-->& ((), v) 
% swap* & (v1, v2) &|-->& (v2, v1) 
% assocl* & (v1, (v2, v3)) &|-->& ((v1, v2), v3) 
% assocr* & ((v1, v2), v3) &|-->& (v1, (v2, v3)) 
% dist & (left v1, v3) &|-->& left (v1, v3)
% dist & (right v2, v3) &|-->& right (v2, v3)
% factor & (left (v1, v3)) &|-->& (left v1, v3) 
% factor & (right (v2, v3)) &|-->& (right v2, v3)   
\end{scriptsize}
The operational semantics of the closure conditions are presented in
the usual big-step style.
\begin{scriptsize}
%subcode{proof} include main
%@ ~
%@@ id v |--> v 
%
%@ c{dagger} v1 |--> v2
%@@ (sym c) v1 |--> v2
%
%@ c1  v1 |--> v
%@ c2  v |--> v2
%@@ (c1(;)c2)  v1 |--> v2
%---
%@ c1  v1 |--> v2
%@@ (c1 (+) c2)  (left v1) |--> left v2
%
%@ c2  v1 |--> v2
%@@ (c1 (+) c2)  (right v1) |--> right v2
%---
%@ c1  v1 |--> v3
%@ c2  v2 |--> v4
%@@ (c1 (*) c2)  (v1, v2) |--> (v3, v4)
\end{scriptsize}

@@

\begin{proposition}[Strong Normalizing]
\label{prop:termination-pi} 
{{Pi}} computations always terminate.  
{{forall. c:b1<->b2, v:b1, exists v':b2.}}  
{{c v |--> v'}}
\end{proposition}

@@

\paragraph*{Booleans, Conditionals and Cloning.} 
We use the type {{1+1}} to denote {{bool}}, with {{left ()}} as
{{true}} and {{right ()}} as {{false}}. The combinator
{{not:bool<->bool}} can be expressed by {{swap+}} at the type
{{bool}}. Further, given any two combinators {{c1:b1<->b2}} and
{{c2:b1<->b2}}, we can write the conditional combinator
{{if_{c1,c2}:bool*b1<->bool*b2}} textually as 
{{dist(;)((id(*)c1)(+)(id(*)c2))(;)factor}} and graphically as:

\dgm{if-c1-c2}

@@

Choosing {{C' =Lft c2 C}}, we have the required derivation sequence: 
{{ <c1+c2, left v1, C> |--> <c1,v1,LeftP c2 C> |-->* {[c1,v1', LeftP c2 C]} |--> {[c1+c2,left v1',C]} }}. The proof follows similarly for the {{right v2}} case.

@@

The syntactic notation above is often obscure and hard to read.
Following the tradition established for monoidal
categories~\cite{springerlink:10.1007/978-3-642-12821-94}, we present
a graphical language that conveys the intuitive semantics of the
language (which is formalized in the next section).

>>>>

For example {{ swap* }} induces a function (akin to)
%subcode{code}
%! language = Haskell
% swapProd (a,b) = (b,a)

The isomorphisms are extended to form a congruence relation by adding the
following constructors that witness equivalence and compatible closure:
%subcode{proof} include main
%@  ~
%@@ id : b <-> b 
%
%@ c : b1 <-> b2
%@@ sym c : b2 <-> b1
%
%@ c1 : b1 <-> b2
%@ c2 : b2 <-> b3
%@@ c1(;)c2 : b1 <-> b3
%---
%@ c1 : b1 <-> b3
%@ c2 : b2 <-> b4
%@@ c1 (+) c2 : b1 + b2 <-> b3 + b4
%
%@ c1 : b1 <-> b3
%@ c2 : b2 <-> b4
%@@ c1 (*) c2 : b1 * b2 <-> b3 * b4

@@

It is important to note that ``values'' and ``isomorphisms'' are 
completely separate syntactic categories which do not intermix.  The 
semantics of the language come when these are made to interact at
the ``top level'', a new syntactic category:
%subcode{bnf} include main
% top-level term, l ::= c v
We refer to this as \emph{application}.

@@

\begin{definition}{(Syntax of {{Pi}})}
\label{def:Pi}
We collect our types, values, and isomorphisms, to get the full language
definition.
%subcode{bnf} include main
% value types, b ::= 0 | 1 | b+b | b*b 
% values, v ::= () | left v | right v | (v,v) 
%
% iso.~types, t ::= b <-> b
% base iso ::= zeroe | zeroi 
%     &|& swap+ | assocl+ | assocr+ 
%     &|& identl* | identr* 
%     &|& swap* | assocl* | assocr* 
%     &|& dist0 | factor0 | dist | factor 
% iso comb., c ::= iso | id | sym c | c (;) c | c (+) c | c (*) c 
% top-level term, l ::= c v
\end{definition}

@@

The operational semantics of top-level terms of {{Pi}} is summarized below
(also see \cite{infeffects}).  The semantics of applying the primitive
combinators to a value is given by the following single-step reductions below.
Since there are no values of type {{0}}, there are no rules for the impossible
cases:
\begin{scriptsize}
%subcode{opsem} include main
%! columnStyle = rlcl
% zeroi & v & |-->& right v
% zeroe & (right v) &|-->& v
% swap+ & (left v) &|-->& right v
% swap+ & (right v) &|-->& left v 
% assocl+ & (left v1) &|-->& left (left v1)
% assocl+ & (right (left v2)) &|-->& left (right v2)
% assocl+ & (right (right v3)) &|-->& right v3 
% assocr+ & (left (left v1)) &|-->& left v1
% assocr+ & (left (right v2)) &|-->& right (left v2)
% assocr+ & (right v3) &|-->& right (right v3)
% identl* & ((), v) &|-->& v 
% identr* & v &|-->& ((), v) 
% swap* & (v1, v2) &|-->& (v2, v1) 
% assocl* & (v1, (v2, v3)) &|-->& ((v1, v2), v3) 
% assocr* & ((v1, v2), v3) &|-->& (v1, (v2, v3)) 
% dist & (left v1, v3) &|-->& left (v1, v3)
% dist & (right v2, v3) &|-->& right (v2, v3)
% factor & (left (v1, v3)) &|-->& (left v1, v3) 
% factor & (right (v2, v3)) &|-->& (right v2, v3)   
\end{scriptsize}
The operational semantics of the closure conditions are presented in
the usual big-step style.
\begin{scriptsize}
%subcode{proof} include main
%@ ~
%@@ id v |--> v 
%
%@ c{dagger} v1 |--> v2
%@@ (sym c) v1 |--> v2
%
%@ c1  v1 |--> v
%@ c2  v |--> v2
%@@ (c1(;)c2)  v1 |--> v2
%---
%@ c1  v1 |--> v2
%@@ (c1 (+) c2)  (left v1) |--> left v2
%
%@ c2  v1 |--> v2
%@@ (c1 (+) c2)  (right v1) |--> right v2
%---
%@ c1  v1 |--> v3
%@ c2  v2 |--> v4
%@@ (c1 (*) c2)  (v1, v2) |--> (v3, v4)
\end{scriptsize}

@@

\begin{proposition}[Strongly Normalizing]
\label{prop:termination-pi} 
{{Pi}} computations always terminate.  
{{forall. c:b1<->b2, v:b1, exists v':b2.}}  
{{c v |--> v'}}
\end{proposition}

@@

\paragraph*{Booleans, Conditionals and Cloning.} 
We use the type {{1+1}} to denote {{bool}}, with {{left ()}} as
{{true}} and {{right ()}} as {{false}}. The combinator
{{not:bool<->bool}} can be expressed by {{swap+}} at the type
{{bool}}. Further, given any two combinators {{c1:b1<->b2}} and
{{c2:b1<->b2}}, we can write the conditional combinator
{{if_{c1,c2}:bool*b1<->bool*b2}} as 
{{dist(;)((id(*)c1)(+)(id(*)c2))(;)factor}}.

% \dgm{if-c1-c2}

@@

Choosing {{C' =LeftP c2 C}}, we have the required derivation sequence: 
{{ <c1+c2, left v1, C> |--> <c1,v1,LeftP c2 C> |-->* {[c1,v1', LeftP c2 C]} |--> {[c1+c2,left v1',C]} }}. The proof follows similarly for the {{right v2}} case.

@@

The syntactic notation above is often obscure and hard to read.
Following the tradition established for monoidal
categories~\cite{springerlink:10.1007/978-3-642-12821-94}, we present
a graphical language that conveys the intuitive semantics of the
language.
